---
title: "Standard-conformant parsing with url_standard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard-conformant parsing with url_standard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rurl)
```

## Why a selector?

`rurl` exposes many low-level normalization knobs (`path_encoding`,
`path_normalization`, `case_handling`, the host model, …). Assembling a coherent
"RFC 3986-conformant" or "WHATWG-conformant" profile by hand means getting
several of them consistent at once, and getting one wrong yields subtly
non-conformant output.

The `url_standard` selector does that assembly for you. It takes one of three
values:

- `NULL` (default) — today's behavior, exactly. Fully backward compatible:
  byte-for-byte identical output and unchanged result shape.
- `"rfc3986"` — the RFC 3986 profile.
- `"whatwg"` — the WHATWG URL Standard profile (on the axes `rurl` governs).

There is **no default flip**: the selector is purely additive, and `NULL`
callers are unaffected. It is accepted by `safe_parse_url()`,
`safe_parse_urls()`, the `get_*()` accessors, `canonical_join()`, and
`resolve_url()`.

## The canonical cases

### Encoded unreserved bytes (`%41%42`)

`%41%42` is the percent-encoding of the ASCII letters `AB` — both *unreserved*
bytes. RFC 3986 §6.2.2.2 says unreserved percent-encodings should be decoded to
their literal form; WHATWG preserves them.

```{r unreserved}
get_path("http://example.com/%41%42", url_standard = "rfc3986")
get_path("http://example.com/%41%42", url_standard = "whatwg")
```

### Encoded reserved bytes (`%2F`)

`%2F` is an encoded `/` — a *reserved* byte. Neither standard decodes it into a
path-separating slash (that would change the path structure), so it stays
encoded under both; only the hex case is canonicalized.

```{r reserved}
get_path("http://example.com/a%2Fb", url_standard = "rfc3986")
get_path("http://example.com/a%2fb", url_standard = "whatwg")
```

### Whole-number IPv4 (`2130706433`)

`2130706433` is the 32-bit integer form of `127.0.0.1`. WHATWG coerces it to a
dotted-decimal address; RFC 3986 has no such rule, so it is treated as a
registered name — but a diagnostic fires under **both** standards so a caller
can tell the host was a numeric shorthand.

```{r ipv4}
get_host("http://2130706433/", url_standard = "whatwg")
get_host_type("http://2130706433/", url_standard = "whatwg")
get_url_diagnostics("http://2130706433/", url_standard = "rfc3986")
```

## Diagnostics are facts, not policy

`url_standard` also unlocks three companion helpers. They return `NA` unless you
pass a selector, and — importantly — they never widen the parse result shape:
metadata is surfaced only through these helpers, never as new columns or fields.

```{r helpers}
get_host_type("http://example.com/", url_standard = "whatwg")
get_scheme_class(c("http://a/", "ftps://a/"), url_standard = "whatwg")
get_url_diagnostics("http://0x7f.1/", url_standard = "whatwg")
```

A diagnostic describes a *shape* of the input; it does not decide what to do
about it. A link-graph builder can ignore `ipv4-*` tokens when computing keys,
while an SSRF/allowlist guard can reject any host that carries one. Because they
are facts, the same token fires under both standards.

## Ports and backslashes

`port_handling` controls whether the port appears in `clean_url`. It is a
standalone editorial knob (independent of `url_standard`), but under `"whatwg"`
its `"keep"` value elides a port that matches its special scheme's default.

```{r ports}
get_clean_url("http://example.com:80/p", port_handling = "keep")
get_clean_url("http://example.com:80/p", port_handling = "keep",
              url_standard = "whatwg")
get_clean_url("http://example.com:8080/p", port_handling = "keep",
              url_standard = "whatwg")
```

Under `"whatwg"`, a literal backslash is recognized as a path separator for the
WHATWG-special schemes (`http`/`https`/`ftp`), as browsers do. `%5C` is never
treated as a separator, and `"rfc3986"` leaves backslashes inert.

```{r backslash}
get_clean_url("http://example.com/a\\b", url_standard = "whatwg")
get_clean_url("http://example.com/a\\b", url_standard = "rfc3986")
```

## Resolving references

`resolve_url()` implements RFC 3986 §5 reference resolution — turning a relative
link and a base URL into an absolute URL — and then canonicalizes the result
with the same machinery, so `url_standard` and the other options flow straight
through.

```{r resolve}
resolve_url("../g", "http://a/b/c/d;p?q")
resolve_url("//other.example/p", "http://a/b/c")
resolve_url(c("g", "../h"), "http://a/b/c/")
```

## Migration notes

- **Pin `url_standard = "whatwg"`** for WHATWG-aligned link identity on the
  governed axes (path percent/dot handling, the host IPv4/reg-name model). This
  collapses a hand-tuned multi-knob profile down to one argument. It is
  deliberately narrower than "browser-faithful": IDNA rendering and query
  handling are not governed by the selector.
- **`path_encoding = "keep"` is a stopgap, not a profile.** If you adopted
  `path_encoding = "keep"` earlier to stop `%2F` reserved-byte false joins, keep
  in mind it is a collision fix, not behavior-equivalent to `"whatwg"`: it
  leaves `%41%42` encoded but does not resolve encoded dot segments and does not
  change host numeric parsing. Prefer the selector once you can.

## What the selector does not govern

The selector governs path percent/dot handling, the host IPv4/reg-name model,
`case_handling`, default-port elision (under `port_handling = "keep"`), and
WHATWG backslash recognition. It does **not** govern IDNA rendering
(`host_encoding`) or query handling (owned by the query options), and it does
not expand the allowed scheme set beyond `http`/`https`/`ftp`/`ftps`.
