---
title: "Standard-conformant parsing with url_standard"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard-conformant parsing with url_standard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(rurl)
```

## Why a selector?

`rurl` exposes many low-level normalization knobs (`path_encoding`,
`path_normalization`, `case_handling`, the host model, …). Assembling a coherent
"RFC 3986-conformant" or "WHATWG-conformant" profile by hand means getting
several of them consistent at once, and getting one wrong yields subtly
non-conformant output.

The `url_standard` selector does that assembly for you. It takes one of three
values:

- `NULL` (default) — today's behavior, exactly. Fully backward compatible:
  byte-for-byte identical output and unchanged result shape.
- `"rfc3986"` — the RFC 3986 profile.
- `"whatwg"` — the WHATWG URL Standard profile (on the axes `rurl` governs).

There is **no default flip**: the selector is purely additive, and `NULL`
callers are unaffected. It is accepted by `safe_parse_url()`,
`safe_parse_urls()`, the `get_*()` accessors, `canonical_join()`, and
`resolve_url()`.

## The canonical cases

### Encoded unreserved bytes (`%41%42`)

`%41%42` is the percent-encoding of the ASCII letters `AB` — both *unreserved*
bytes. RFC 3986 §6.2.2.2 says unreserved percent-encodings should be decoded to
their literal form; WHATWG preserves them.

```{r unreserved}
get_path("http://example.com/%41%42", url_standard = "rfc3986")
get_path("http://example.com/%41%42", url_standard = "whatwg")
```

### Encoded reserved bytes (`%2F`)

`%2F` is an encoded `/` — a *reserved* byte. Neither standard decodes it into a
path-separating slash (that would change the path structure), so it stays
encoded under both; only the hex case is canonicalized.

```{r reserved}
get_path("http://example.com/a%2Fb", url_standard = "rfc3986")
get_path("http://example.com/a%2fb", url_standard = "whatwg")
```

### Whole-number IPv4 (`2130706433`)

`2130706433` is the 32-bit integer form of `127.0.0.1`. WHATWG coerces it to a
dotted-decimal address; RFC 3986 has no such rule, so it is treated as a
registered name — but a diagnostic fires under **both** standards so a caller
can tell the host was a numeric shorthand.

```{r ipv4}
get_host("http://2130706433/", url_standard = "whatwg")
get_host_type("http://2130706433/", url_standard = "whatwg")
get_url_diagnostics("http://2130706433/", url_standard = "rfc3986")
```

## Diagnostics are facts, not policy

`url_standard` also unlocks three companion helpers. They return `NA` unless you
pass a selector, and — importantly — they never widen the parse result shape:
metadata is surfaced only through these helpers, never as new columns or fields.

```{r helpers}
get_host_type("http://example.com/", url_standard = "whatwg")
get_scheme_class(c("http://a/", "ftps://a/"), url_standard = "whatwg")
get_url_diagnostics("http://0x7f.1/", url_standard = "whatwg")
```

A diagnostic describes a *shape* of the input; it does not decide what to do
about it. A link-graph builder can ignore `ipv4-*` tokens when computing keys,
while an SSRF/allowlist guard can reject any host that carries one. Because they
are facts, the same token fires under both standards.

## Ports and backslashes

`port_handling` controls whether the port appears in `clean_url`. It is a
standalone editorial knob (independent of `url_standard`), but under `"whatwg"`
its `"keep"` value elides a port that matches its special scheme's default.

```{r ports}
get_clean_url("http://example.com:80/p", port_handling = "keep")
get_clean_url("http://example.com:80/p", port_handling = "keep",
              url_standard = "whatwg")
get_clean_url("http://example.com:8080/p", port_handling = "keep",
              url_standard = "whatwg")
```

Under `"whatwg"`, a literal backslash is recognized as a path separator for the
WHATWG-special schemes (`http`/`https`/`ftp`), as browsers do. `%5C` is never
treated as a separator, and `"rfc3986"` leaves backslashes inert.

```{r backslash}
get_clean_url("http://example.com/a\\b", url_standard = "whatwg")
get_clean_url("http://example.com/a\\b", url_standard = "rfc3986")
```

## Resolving references

`resolve_url()` turns a relative link and a base URL into an absolute URL, then
canonicalizes the result with the same machinery, so `url_standard` and the
other options flow straight through.

Resolution itself is one of the axes the selector governs. Under `"rfc3986"` —
and under the default `NULL` — the merge is RFC 3986 §5 exactly.

```{r resolve}
resolve_url("../g", "http://a/b/c/d;p?q")
resolve_url("//other.example/p", "http://a/b/c")
resolve_url(c("g", "../h"), "http://a/b/c/")
```

### What `"whatwg"` changes

Under `"whatwg"` the WHATWG URL Standard's reference-*parsing* rules run before
the merge. Four of them are visible:

```{r resolve-whatwg}
# 1. A reference carrying the base's OWN special scheme is relative, not
#    absolute -- so this is not read as the absolute `http:g`.
resolve_url("http:g", "http://a/b/c/d", url_standard = "whatwg")

# 2. Under a special base, `\` reads as `/`.
resolve_url("..\\g", "http://a/b/c/d", url_standard = "whatwg")

# 3. A leading run of slashes-or-backslashes introduces an authority.
resolve_url("///g", "http://a/b/c/d", url_standard = "whatwg")

# 4. Leading and trailing C0-or-space is stripped before the reference is read.
resolve_url("  \n g ", "http://a/b/c/d", url_standard = "whatwg")
```

### Which colon starts a scheme

One rule is not WHATWG's — it is the RFC's own grammar, so it applies under
**both** named selectors. Deciding whether a reference is absolute means
deciding whether the text before its first colon is a scheme, and RFC 3986
§3.1's production is `ALPHA *( ALPHA / DIGIT / "+" / "-" / "." )`. The looser
group in the RFC's Appendix B parser, which the appendix itself describes as
non-validating, admits things that are not schemes — and §4.2 names the
consequence: a relative path that merely *contains* a colon gets read as
absolute and the base is thrown away.

```{r resolve-scheme}
# `10.0.0.7` cannot be a scheme, so this is a relative path.
resolve_url("10.0.0.7:8080/foo.html", "file:///some/dir/bar.html",
            url_standard = "rfc3986")

# The same tightening cuts the other way for a BASE: a base whose scheme
# existed only under the loose group is not an absolute URL, so there is
# nothing to resolve against.
resolve_url("g", "10.0.0.7:8080/dir/x", url_standard = "rfc3986")
```

`url_standard = NULL` keeps the loose group. Tightening it there would move
frozen bytes rather than merely improve them, and the freeze governs whether
output may *move*, not whether it is right — so the fix ships on the two
selectors that claim a standard, and the one that claims none is left alone.

```{r resolve-null-frozen}
resolve_url("g", "10.0.0.7:8080/dir/x")
```

### Asking for the standard string instead

`resolve_url()` returns `clean_url` by default, which drops credentials and the
fragment — a cleaning product, not a standard serialization. When you need the
surface RFC 3986 §5.4's own worked examples are reproducible on, ask for it:

```{r resolve-serialized}
resolve_url("../g", "http://u:pw@a/b/c/d?q#f", url_standard = "whatwg")
resolve_url("../g", "http://u:pw@a/b/c/d?q#f", url_standard = "whatwg",
            output = "serialized")
```

`output = "serialized"` requires an explicit `url_standard`, since `NULL`
selects no standard to serialize *to*.

## Migration notes

- **Pin `url_standard = "whatwg"`** for WHATWG-aligned link identity on the
  governed axes (path percent/dot handling, the host IPv4/reg-name model). This
  collapses a hand-tuned multi-knob profile down to one argument. It is
  deliberately narrower than "browser-faithful": IDNA rendering and query
  handling are not governed by the selector.
- **`path_encoding = "keep"` is a stopgap, not a profile.** If you adopted
  `path_encoding = "keep"` earlier to stop `%2F` reserved-byte false joins, keep
  in mind it is a collision fix, not behavior-equivalent to `"whatwg"`: it
  leaves `%41%42` encoded but does not resolve encoded dot segments and does not
  change host numeric parsing. Prefer the selector once you can.

## What the selector does not govern

The selector governs path percent/dot handling, the host IPv4/reg-name model,
`case_handling`, default-port elision (under `port_handling = "keep"`), and
WHATWG backslash recognition. It does **not** govern IDNA rendering
(`host_encoding`) or query handling (owned by the query options).

It also does not decide **which schemes get parsed at all** — that is a separate
axis, `scheme_acceptance`. The default `"web"` admits only the curated
`http`/`https`/`ftp`/`ftps`/`file` allowlist; `"general"` additionally accepts
non-special schemes, including opaque-path ones such as `mailto:`, `data:` and
`tel:`. The two axes compose: `scheme_acceptance` decides what gets parsed,
`url_standard` decides how the result is interpreted.

```{r acceptance}
# Rejected by the default "web" allowlist ...
get_host("mailto:jane@example.com", url_standard = "whatwg")

# ... parsed under "general".
get_host("mailto:jane@example.com", url_standard = "whatwg",
         scheme_acceptance = "general")
```

A `mailto:` URL is a WHATWG **opaque path**, which by definition has no
authority, so the parse result reports none — the recipient stays in `path`. The
accessor above is doing something different and deliberate: extracting the
web-y facts *about* an address. The two answers are allowed to differ.

```{r opaque}
safe_parse_urls("mailto:jane@example.com", url_standard = "whatwg",
                scheme_acceptance = "general")[, c("host", "user", "path")]
```

## Conformance posture

`rurl`'s conformance claims are measured, frozen and regenerable — the studies
live in [`analysis/`](https://gitlab.com/bart-turczynski/rurl/-/tree/main/analysis),
with the harnesses in `inst/bench/`.

**WHATWG.** Against the spec's own conformance suite (web-platform-tests
`urltestdata.json`), at `scheme_acceptance = "general"` the `"whatwg"` profile
accepts all **336** success cases with full component parity — scheme,
username, password, host, port, path, query and fragment — and correctly
rejects all **202** failure cases: **538/538**. At the default `scheme_acceptance = "web"` the same corpus
scores 176 accepted, all at full parity, plus the same 202 rejections; the 160
difference is the ADR 0004 allowlist declining non-web schemes before the
grammar is consulted, not a conformance miss. Against `adaR` (the
Ada C++ WHATWG reference) over a 336-input adversarial corpus, it differs on
**two rows**, and neither is a parsing disagreement: one is punycode-vs-Unicode
*presentation* (ADR 0002 — reachable through `host_encoding`), and one is the
held `scheme_policy = "require"` axis. On the rows where both parsers accept,
scheme, port and path agreement are all 1.000.

**RFC 3986.** Over the 257-row audited conformance fixture, the `"rfc3986"`
profile matches the standard on **164** rows and departs on **93** — 81 where it
rejects what the RFC admits (all of them the ADR 0004 host/authority gate; the
closed scheme set contributes none) and 12 where it accepts what the RFC does not (ADR 0002 Unicode
hosts, ADR 0011 readable paths). Every one of the 93 cites the ADR or ticket
that owns it; none is accidental. A hand-authored, two-sided probe set covering
the grammar and normalization rules passes 32/32 in both directions, with five
deliberate departures pinned *outside* the score so strictness cannot inflate a
conformance number.

Two boundaries keep that honest rather than triumphal:

- The WPT fixture behind that 336/336 keeps only **absolute** URLs (rows with a
  null base), so the figure says **nothing** about base-relative resolution. It
  does now cover every scheme the suite exercises, including opaque, `ws:` and
  `wss:`. Base-relative resolution is measured **separately**, over the
  complementary 274 base-carrying rows of the same upstream file: **247 exact,
  27 enumerated differences** (18 Windows drive letters under `file:`, 5 a
  recomposition seam, 4 absolute references whose deviation is in absolute
  parsing). **The two are disjoint populations and are never added together** —
  there is no combined rate, and the 247/274 split is a known-differ set for
  regression tracking, not a conformance rate.
- The `"whatwg"` profile is WHATWG on the axes it governs — **not** a full
  UTS-46 host mapping. Ligatures, circled digits and zero-width code points pass
  through unmapped (ADR 0002).

The residual WHATWG deviations are both dials rather than defects: each is
reachable from a documented argument, which is the design's whole claim.
